---
title: "Hands-on Exercise 1"
author: "yang yayong"
---

# 1  A Layered Grammar of Graphics: ggplot2 methods

## 1.1 Learning Outcome

In this chapter, we will learn the basic principles and essential components of ggplot2. At the same time, we will gain hands-on experience on using these components to plot statistical graphics based on the principle of Layered Grammar of Graphics. By then end of this chapter we will be able to apply the essential graphical elements provided by ggplot2 to create elegant and yet functional statistical graphics.

## 1.2 Getting started

### 1.2.1 Installing and loading the required libraries

```{r}
pacman::p_load(tidyverse)
```

### 1.2.2 Importing data

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

## 1.3 Introducing ggplot

## 1.3.1 R Graphics VS ggplot

::: panel-tabset
## R Graphics

```{r}
hist(exam_data$MATHS)
```

## ggplot2

```{r}
ggplot(data=exam_data, aes(x = MATHS)) +
  geom_histogram(bins=10, 
                 boundary = 100,
                 color="black", 
                 fill="grey") +
  ggtitle("Distribution of Maths scores")
```
:::

## 1.4 Grammar of Graphics

### 1.4.1 A Layered Grammar of Graphics

A short description of each building block are as follows:

-   **Data**: The dataset being plotted.

-   **Aesthetics** take attributes of the data and use them to influence visual characteristics, such as position, colours, size, shape, or transparency.

-   **Geometrics**: The visual elements used for our data, such as point, bar or line.

-   **Facets** split the data into subsets to create multiple variations of the same graph (paneling, multiple plots).

-   **Statistics**, statiscal transformations that summarise data (e.g. mean, confidence intervals).

-   **Coordinate systems** define the plane on which data are mapped on the graphic.

-   **Themes** modify all non-data components of a plot, such as main title, sub-title, y-aixs title, or legend background.

## 1.5 Essential Grammatical Elements in ggplot2: data

Let us call the `ggplot()` function using the code chunk on the right.

```{r}
ggplot(data=exam_data)
```

::: note
-   A blank canvas appears.
-   `ggplot()` initializes a ggplot object.
-   The `data` argument defines the dataset to be used for plotting.
-   If the dataset is not already a `data.frame`, it will be converted to one by `fortify()`.
:::

## 1.6 Essential Grammatical Elements in ggplot2:

## [Aesthetic mappings](https://ggplot2.tidyverse.org/articles/ggplot2-specs.html)

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS))
```

::: Note
-   ggplot includes the x-axis and the axis’s label.
:::

## 1.7 Essential Grammatical Elements in ggplot2: geom

Geometric objects are the actual marks we put on a plot. Examples include:

-   *geom_point* for drawing individual points (e.g., a scatter plot)

-   *geom_line* for drawing lines (e.g., for a line charts)

-   *geom_smooth* for drawing smoothed lines (e.g., for simple trends or approximations)

-   *geom_bar* for drawing bars (e.g., for bar charts)

-   *geom_histogram* for drawing binned values (e.g. a histogram)

-   *geom_polygon* for drawing arbitrary shapes

-   *geom_map* for drawing polygons in the shape of a map! (You can access the data to use for these maps by using the map_data() function).

### 1.7.1 Geometric Objects: geom_bar

The code chunk below plots a bar chart by using [`geom_bar()`](https://ggplot2.tidyverse.org/reference/geom_bar.html).

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()
```

### 1.7.2 Geometric Objects: geom_dotplot

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(dotsize = 0.5)
```

::: Be warned The y scale is not very useful, in fact it is very misleading. :::

::: note
The code chunk below performs the following two steps:

-   scale_y_continuous() is used to turn off the y-axis, and

-   binwidth argument is used to change the binwidth to 2.5.
:::

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(binwidth=2.5,         
               dotsize = 0.5) +      
  scale_y_continuous(NULL,           
                     breaks = NULL)  
```

### 1.7.3 Geometric Objects: `geom_histogram()`

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_histogram()       
```

::: Note
Note that the default bin is 30.
:::

### 1.7.4 Modifying a geometric object by changing geom()

In the code chunk below,

-   bins argument is used to change the number of bins to 20,

-   fill argument is used to shade the histogram with light blue color, and

-   color argument is used to change the outline colour of the bars in black

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20,            
                 color="black",      
                 fill="light blue")  
```

### 1.7.5 Modifying a geometric object by changing *aes()*

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           fill = GENDER)) +
  geom_histogram(bins=20, 
                 color="grey30")
```

::: Note
This approach can be used to colour, fill and alpha of the geometric.
:::

1.7.6 Geometric Objects: geom-density()

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_density()           
```

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS, 
           colour = GENDER)) +
  geom_density()
```

### 1.7.7 Geometric Objects: geom_boxplot

[`geom_boxplot()`](https://ggplot2.tidyverse.org/reference/geom_boxplot.html) displays continuous value list. It visualises five summary statistics (the median, two hinges and two whiskers), and all “outlying” points individually.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS,       
           x= GENDER)) +    
  geom_boxplot()            
```

[**Notches**](https://sites.google.com/site/davidsstatistics/home/notched-box-plots) are used in box plots to help visually assess whether the medians of distributions differ. If the notches do not overlap, this is evidence that the medians are different.

The code chunk below plots the distribution of Maths scores by gender in notched plot instead of boxplot.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot(notch=TRUE)
```

### 1.7.8 Geometric Objects: geom_violin

[`geom_violin`](https://ggplot2.tidyverse.org/reference/geom_violin.html) is designed for creating violin plot. Violin plots are a way of comparing multiple data distributions. With ordinary density curves, it is difficult to compare more than just a few distributions because the lines visually interfere with each other. With a violin plot, it’s easier to compare several distributions since they’re placed side by side.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_violin()
```

### 1.7.9 Geometric Objects: geom_point()

[`geom_point()`](https://ggplot2.tidyverse.org/reference/geom_point.html) is especially useful for creating scatterplot.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point()            
```

### 1.7.10 *geom* objects can be combined

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot() +                    
  geom_point(position="jitter", 
             size = 0.5)        
```

## 1.8 Essential Grammatical Elements in ggplot2: stat

### 1.8.1 Working with `stat()`

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot()
```

### 1.8.2 Working with stat - the *stat_summary()* method

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  stat_summary(geom = "point",       
               fun = "mean",         
               colour ="red",        
               size=4)               
```

### 1.8.3 Working with stat - the `geom()` method

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  geom_point(stat="summary",        
             fun="mean",           
             colour="red",          
             size=4)          
```

### 1.8.4 Adding a best fit curve on a scatterplot?

```{r}
#| echo: false
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point()            
```

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(size=0.5)
```

::: Note
The default method used is loess.
:::

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              linewidth=0.5)
```

## 1.9 Essential Grammatical Elements in ggplot2: Facets

### 1.9.1 Working with `facet_wrap()`

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_wrap(~ CLASS)
```

### 1.9.2 `facet_grid()` function

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_grid(~ CLASS)
```

## 1.10 Essential Grammatical Elements in ggplot2: Coordinates

### 1.10.1 Working with Coordinate

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()
```

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip()
```

### 1.10.2 Changing the y- and x-axis range

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, size=0.5)
```

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
```

## 1.11 Essential Grammatical Elements in ggplot2: themes

### 1.11.1 Working with theme

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_gray()
```

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_classic()
```

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_minimal()
```

## 1.12 Reference

-   Hadley Wickham (2023) [**ggplot2: Elegant Graphics for Data Analysis**](https://ggplot2-book.org/). Online 3rd edition.

-   Winston Chang (2013) [**R Graphics Cookbook 2nd edition**](https://r-graphics.org/). Online version.

-   Healy, Kieran (2019) [**Data Visualization: A practical introduction**](https://socviz.co/). Online version

-   [Learning ggplot2 on Paper – Components](https://henrywang.nl/learning-ggplot2-on-paper-components/)

-   [Learning ggplot2 on Paper – Layer](https://henrywang.nl/learning-ggplot2-on-paper-layer/)

-   [Learning ggplot2 on Paper – Scale](https://henrywang.nl/tag/learning-ggplot2-on-paper/)
