---
title: "In-class Ex4"
author: "yang yayong"
date: "May 10 2025"
execute:
  eval: true
  echo: true
  message: false
  freeze: true
  warning: FALSE
---

The `ggstatsplot` package offers a convenient way to generate various types of plots by adjusting a few parameters.

Basic plots are effective for exploratory analysis and initial insights, whereas statistically enriched plots are valuable for drawing conclusions and supporting scientific reasoning.

```{r}
pacman::p_load(ggstatsplot, tidyverse)
```

```{r}
exam <- read_csv("data/Exam_data.csv")
```

### Two-sample mean test: *ggbetweenstats()*

For example changing the type from

np : This one is showing median.

p : This will show mean value.

r : This plot has remove the outliers.

```{r}
ggbetweenstats(
  data = exam,
  x = GENDER, 
  y = MATHS,
  type = "np",
  messages = FALSE
)
```

```{r}
ggbetweenstats(
  data = exam,
  x = GENDER, 
  y = MATHS,
  type = "p",
  messages = FALSE
)
```

```{r}
ggbetweenstats(
  data = exam,
  x = GENDER, 
  y = MATHS,
  type = "r",
  messages = FALSE
)
```

::: callout-note
If you need show the statistic information,it would better advise to use ggstatsplot instead of ggplot2.
:::

The difference between marginal = FALSE and marginal = TRUE

```{r}
ggscatterstats(
  data = exam,
  x = MATHS,
  y = ENGLISH,
  marginal = FALSE,
  )
```

```{R}
ggscatterstats(
  data = exam,
  x = MATHS,
  y = ENGLISH,
  marginal = TRUE,
  )
```

Compare to what we learn from DAL mosaic plot and the plot belowed:

```{r}
exam1 <- exam %>% 
  mutate(MATHS_bins = 
           cut(MATHS, 
               breaks = c(0,60,75,85,100))
)

```

```{r}
ggbarstats(exam1, 
           x = MATHS_bins, 
           y = GENDER)
```

The one can reflect the insights for Significant Test of Association within the target group
